Abstractions for Multi-Sorted Substitutions

### 2023-09-18 paper/paper.tex
Update.

  Similarly, for a typed language formalized via extrinsic typing we
  need type preservation lemmas for each substitution and renaming
  operation, which again doubles the repetition.

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization with three sorts of variables
  can easily end up with 24 definitions for each operation related
  to substitution, which usually includes at least a substitution type, and
  operations for application, extension, weakening, lifting, singleton
  substitution, extensional equality, and an identity substitution.

  In summary, formalizing a language with expression-, type-, and
  kind-variables with a naive approach can end up with a total of 24
  repetitions for each definition related to substitution, not counting
  the interaction lemmas between different substitutions that are
  acting on the same syntactic categories!

  In our framework we address those repetitions as followed:
  \begin{itemize}
  \item
    Instead of requiring a separate substitution for each sort of variable,
    we define a multi-sorted substitution which replaces all variables
    of all sorts simultaneously and can be applied to terms of all syntactic categories.
    This not only unifies many different definitions, removing redundancy on the way,
    but also eliminates the need for interaction lemmas.
  \item
    We use de Bruijn indices, but unify renamings and typings using the Kit abstraction.
    We further developed a Kit-like abstraction to also unify
    compositions of renamings and substitutions.
  \end{itemize}

  A close examination of the formalizations reveals that variable
  substitution behaves uniformly across languages, in the sense that
  the behavior of substitution is completely determined by the positions
  of variable bindings and variable uses.

  Formalizing a typed programming language in a proof assistant
  requires to choose between an extrinsic and intrinsic representation
  of typing.

  With \emph{intrinsic typing}, the definitions of syntax and typing
  are merged into a single definition of well-typed terms.
  Consequently, the definition of an operation on terms also acts
  as a proof of type preservation for that operation, as it has to
  map well-typed terms to well-typed terms.
  For example, the definition of substitution
  also acts as a proof that substitution preserves typing, and the
  definition of a small-step semantics also acts as a proof for subject
  reduction.
  This often has the benefit of more concise code, as shared structure
  between definitions and preservation lemmas does not need to be
  repeated.

  With \emph{extrinsic typing}, the definitions of syntax and typing
  are kept separately: the syntax is defined first, and then typing is
  defined as a relation. Consequently, operations on terms and
  their type preservation lemmas are also separate entities.
  This has the benefit that specification and proofs are kept strictly separate,
  allowing a human to judge whether a specification is correct without also
  having to understand the proofs.
  % To understand the specification of intrinsic typing one has to
  % additionally understand the proofs, which can introduce significant
  % cognitive overhead.

  For our substition framework, we choose to work with extrinsic
  typing for a different reason.

  % For substitution, this separation brings an opportunity for automation:
  % free from the arbitrary complexity of typing relations,
  % the definitions and lemmas of substitution for untyped terms
  % behaves uniformly across object languages, and can be derived
  % completely.

  % While extrinsic typing cannot eliminate the shared structure between
  % definitions and preservation lemmas, it can automated by reflection
  % or generic programming techniques, leaving only the preservation lemmas
  % to be proven by the programmer. While typing can be arbitrary complex,
  % the definition and metatheory of untyped substitution behaves the same
  % across languages.

  % However, when working with library support, this is less of a benefit.
  % Most operations, like substitution, have a generic definition and
  % metatheory across languages, which can be derived automatically via
  % reflection or generic programming techniques.
  % Only by adding a typing-relation

  % However, intrinsic typing also has three major drawbacks:
  % \begin{enumerate}
  % \item
  %   It is more difficult to judge whether a specification is correct,
  %   because there is no clear separation between specification and proofs.
  %   To understand the specification one has to additionally
  %   understand the proofs, which can introduce significant cognitive overhead.
  %   % For example specifying a small step semantics is directly intertwined
  %   % with the proof of subject reduction.
  % \item
  %   It is less uniform for object languages with multiple variable
  %   sorts, e.g. type- and expression-variables. In such a scenario,
  %   types are usually indexed by kinds and expressions by kinds and types.
  %   This requires not only a type-in-type and expression-in-expression
  %   substitution, but also a type-in-expression substitution, which causes
  %   interactions between different types of substitutions introducing
  %   additional complexity.
  %   The duplication increases even more if one additionally introduces
  %   kind-variables. Then a total of 6 different substitutions and
  %   their interactions is required.
  %   Unifying these substitutions to reduce repetition is difficult
  %   with intrinsic typing due to the interdependence of definitions.
  % \item
  %   It is less suitable for automation, because type preservation is baked
  %   into the syntax definition and type preservation can become arbitrary complex
  %   depending on the typing relation. This can prevent substitition
  %   definition and lemmas from being derived, where otherwise it would be possible.
  % \end{enumerate}
  % With extrinsic typing the major drawback is the duplication between
  % definitions and type preservation lemmas, which we found more
  % suitable to be eliminated by automation.

  % To combat those drawbacks, we went in the direction of intrinsically
  % scoped syntax in our search for a general, clean, and automated framework
  % for metatheory.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases they are formulated for intrinsic typing.

@@

>>>>

  Similarly, for a typed language formalized via extrinsic typing we
  need type preservation lemmas for each substitution and renaming
  operation, which again doubles the number of definitions.

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization of a language with type and
  kind polymorphism can easily end up with 24 slightly changed copies of
  the whole substitution machinery, which for de Bruijn indices tends to
  include at least a type for parallel substitution, and operations for
  application, extension, weakening, lifting, singleton substitution,
  extensional equality, and an identity substitution - and this is not
  counting the slightly changed copies of interaction lemmas!

  Clearly, this situation is in need of automation.
  Our framework approaches this problem using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  using a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.
  Compared to many other approaches, we do not derive substitutions
  for each of the variable sorts and syntactic categories, but instead
  use a novel approach for defining syntax, which directly supports
  substitutions that replace all variable sorts in parallel and can be
  applied to all syntactic categories.
  By further unifying renamings and substitutions, we gain the ability
  to talk abstractly about any kind of renaming or substitution that
  can occur in the formalization.
  Together with abstractions to unify renamings and substitutions,
  this allows us to create an abstraction that is general enough to
  prove type preservation for all sub

  % Choosing an extrinsic typing approach might seem counter intuitive,
  % as separating

  % In our framework we address those repetitions as followed:
  % \begin{itemize}
  % \item
  %   Instead of requiring a separate substitution for each sort of variable,
  %   we define a multi-sorted substitution which replaces all variables
  %   of all sorts simultaneously and can be applied to terms of all syntactic categories.
  %   This not only unifies many different definitions, removing redundancy on the way,
  %   but also eliminates the need for interaction lemmas.
  % \item
  %   We use de Bruijn indices, but unify renamings and typings using the Kit abstraction.
  %   We further developed a Kit-like abstraction to also unify
  %   compositions of renamings and substitutions.
  % \end{itemize}

  % A close examination of the formalizations reveals that variable
  % substitution behaves uniformly across languages, in the sense that
  % the behavior of substitution is completely determined by the positions
  % of variable bindings and variable uses.

  % Formalizing a typed programming language in a proof assistant
  % requires to choose between an extrinsic and intrinsic representation
  % of typing.

  % With \emph{intrinsic typing}, the definitions of syntax and typing
  % are merged into a single definition of well-typed terms.
  % Consequently, the definition of an operation on terms also acts
  % as a proof of type preservation for that operation, as it has to
  % map well-typed terms to well-typed terms.
  % For example, the definition of substitution
  % also acts as a proof that substitution preserves typing, and the
  % definition of a small-step semantics also acts as a proof for subject
  % reduction.
  % This often has the benefit of more concise code, as shared structure
  % between definitions and preservation lemmas does not need to be
  % repeated.

  % With \emph{extrinsic typing}, the definitions of syntax and typing
  % are kept separately: the syntax is defined first, and then typing is
  % defined as a relation. Consequently, operations on terms and
  % their type preservation lemmas are also separate entities.
  % This has the benefit that specification and proofs are kept strictly separate,
  % allowing a human to judge whether a specification is correct without also
  % having to understand the proofs.
  % % To understand the specification of intrinsic typing one has to
  % % additionally understand the proofs, which can introduce significant
  % % cognitive overhead.

  % For our substition framework, we choose to work with extrinsic
  % typing for a different reason.

  % % For substitution, this separation brings an opportunity for automation:
  % % free from the arbitrary complexity of typing relations,
  % % the definitions and lemmas of substitution for untyped terms
  % % behaves uniformly across object languages, and can be derived
  % % completely.

  % % While extrinsic typing cannot eliminate the shared structure between
  % % definitions and preservation lemmas, it can automated by reflection
  % % or generic programming techniques, leaving only the preservation lemmas
  % % to be proven by the programmer. While typing can be arbitrary complex,
  % % the definition and metatheory of untyped substitution behaves the same
  % % across languages.

  % % However, when working with library support, this is less of a benefit.
  % % Most operations, like substitution, have a generic definition and
  % % metatheory across languages, which can be derived automatically via
  % % reflection or generic programming techniques.
  % % Only by adding a typing-relation

  % % However, intrinsic typing also has three major drawbacks:
  % % \begin{enumerate}
  % % \item
  % %   It is more difficult to judge whether a specification is correct,
  % %   because there is no clear separation between specification and proofs.
  % %   To understand the specification one has to additionally
  % %   understand the proofs, which can introduce significant cognitive overhead.
  % %   % For example specifying a small step semantics is directly intertwined
  % %   % with the proof of subject reduction.
  % % \item
  % %   It is less uniform for object languages with multiple variable
  % %   sorts, e.g. type- and expression-variables. In such a scenario,
  % %   types are usually indexed by kinds and expressions by kinds and types.
  % %   This requires not only a type-in-type and expression-in-expression
  % %   substitution, but also a type-in-expression substitution, which causes
  % %   interactions between different types of substitutions introducing
  % %   additional complexity.
  % %   The duplication increases even more if one additionally introduces
  % %   kind-variables. Then a total of 6 different substitutions and
  % %   their interactions is required.
  % %   Unifying these substitutions to reduce repetition is difficult
  % %   with intrinsic typing due to the interdependence of definitions.
  % % \item
  % %   It is less suitable for automation, because type preservation is baked
  % %   into the syntax definition and type preservation can become arbitrary complex
  % %   depending on the typing relation. This can prevent substitition
  % %   definition and lemmas from being derived, where otherwise it would be possible.
  % % \end{enumerate}
  % % With extrinsic typing the major drawback is the duplication between
  % % definitions and type preservation lemmas, which we found more
  % % suitable to be eliminated by automation.

  % % To combat those drawbacks, we went in the direction of intrinsically
  % % scoped syntax in our search for a general, clean, and automated framework
  % % for metatheory.

@@

  Frameworks for programming languages metatheory can be broadly
  categorized by representation of variable bindings, representation
  of typing and scoping, and requirements on the meta-language.

  Variable bindings have been represented as de Bruijn indices, de
  Bruijn levels, names, locally nameless, locally named, shifted names,
  nominal techniques, (parameterized) higher order abstract syntax.

  Typing and scoping can be both extrinsic, both intrinsic, or
  extrinsically typed and intrinsically scoped.

  As the amount of related work is rather large, we focus on work that
  is more closely related to ours and refer to other papers for the
  broader picture. In particular we cover abstractions for renamings
  and substitutions, comparable extrinsically typed approaches, but also
  intrinsically typed approaches that could be used as a basis for
  multi-sorted syntax, which as mentioned before is intrinsically typed
  in disguise.

  \subsection{Unifying Renamings \& Substitution}

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases kits are formulated for intrinsic typing
  and scenarios with polymorphism are not considered.

  \subsection{Extrinsically Typed Approaches}

@@

  \subsection{Intrinsically Typed Approaches}
